A Journey from Wild to Textbook Data to Reproducibly Refresh the Wages Data from the National Longitudinal Survey of Youth Database

### 2022-05-04 paper/paper.Rmd
Adress minor comments.

---
title: "A Journey from Wild to Textbook Data to Reproducibly Refresh the Wages Data from the National Longitudinal Survey of Youth Database"
authors:
  - name: Dewi Amaliah
    thanks: ""
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: dlamaleeah@gmail.com
  - name: Dianne Cook
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: dicook@monash.edu
  - name: Emi Tanaka
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: emi.tanaka@monash.edu
  - name: Kate Hyde
    thanks: ""
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: "hyde.kate.a@gmail.com"
  - name: Nicholas Tierney
    thanks: ""
    department: ""
    affiliation: Telethon Kids Institute
    location: Nedlands, WA 6009
    email: "nicholas.tierney@gmail.com"
bibliography: references.bib
biblio-style: unsrt
#preamble: >
header-includes:
  - \usepackage{tcolorbox}
  - \usepackage{fontawesome}
  - \usepackage{color, colortbl}
output: 
  #rticles::arxiv_article:
    #keep_tex: false
  bookdown::pdf_book:
    base_format: rticles::arxiv_article
date: "24/02/2021"
always_allow_html: true
keywords: "Data cleaning; Data tidying; Reproducible workflow; Longitudinal data; NLSY79; Initial data analysis;"
abstract: > 
 Textbook data is essential for teaching statistics and data science methods because they are clean, allowing the instructor to  focus on methodology. Ideally textbook data sets are refreshed regularly, especially when they are subsets taken from an on-going data collection. It is also important to use contemporary data for teaching, to imbue the sense that the methodology is relevant today. This paper describes the trials and tribulations of refreshing a textbook data set on wages, extracted from the National Longitudinal Survey of Youth (NLSY79) in the early 1990s. The data is useful for teaching modeling and exploratory analysis of longitudinal data. Subsets of NLSY79, including the wages data, can be found in supplementary files from numerous textbooks and research articles. The NLSY79 database has been continuously updated through to 2018, so new records are available. Here we describe our journey to re-create the wages data, and document the process so that the data can be regularly updated into the future. Our journey was difficult because the steps and decisions taken to get from the raw data to the wages textbook subset, have not been clearly articulated. We have been diligent to provide a reproducible workflow for others to follow, which also hopefully inspires more attempts at refreshing data for teaching. Three new data sets and the code to produce them are provided in the open source R package, called [CENSORED]. 
---

@@

Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques. An example of this is the wages data made public by @SingerJudithD2003Alda in their book, "Applied longitudinal data analysis",  which can be used to teach generalized linear models, in addition to hierarchical, mixed effects, and multilevel models. The data records hourly wages of a sample of high school dropouts, from 1979-1994, along with the demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) [@nlsy79].

The story from modeling the data (and as reported by Singer and Willett) is that wages increase with the length of time in the workforce, a higher level of education leads to higher wages, and that race makes a difference, on average. An exploratory analysis reveals, however, that the individual experience varies a great deal from the overall average. Some individuals experience a decline in wages the longer they are in the workforce and many experience volatility in their wages. The wages data was used to illustrate exploratory longitudinal data analysis in @ilk2004, and was further developed into a case study for use in the teaching of exploratory data analysis at `r censor_word("Iowa State University")`.

@@

This disparity between the average and the individual is a part of statistics, as a discipline that requires more attention. This particular data set is a prime example of discussing this disparity. Textbook data sets have longevity if there is an unresolved mystery. The iris data [@iris-data] is a prime example. It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis. (A side note: the iris data is best replaced today with the penguins data [@penguins-data], which has similar qualities, is new, and does not suffer from a connection with eugenics [@notiris]. We argue that the wages data is in this class of textbook data, too, because it presents a challenge for longitudinal data analysis: how can we better summarize and explain the individual experience?

For the field of statistics and data science by association, it is increasingly important to reach the individual. One might describe this as a divergence of purpose, statistics for public policy, or statistics for the public. The two are not the same. As the world becomes more electronically connected, combating misinformation and mitigating conspiracy theories require that statistics address the individual. For example, with the wages data, even though the message for public policy is that demographic profile is related to different wage patterns on average, the message for the individual is that you are more than your demographic. The majority of people in the study do not have a pattern that is similar to the average. If you have a bad experience, that your wages have declined over time, you are not alone; there are others like you, and more than you think. A similar tone is echoed occasionally in the public media. For example, an article published in the Sydney Morning Herald argues that there is no average Australian [@notaverage].

@@

This paper demonstrates the steps of cleaning data, including subjective decisions made on dealing with anomalies, and documents the process, as recommended by @HuebnerMariannePhD2016Asat. They emphasize that making the data cleaning process accountable and transparent is imperative and essential for the integrity of downstream statistical analyses and model building. Clean data often then goes through an "initial data analysis" (IDA) [@Chatfield1985TIEo], where one would summarize and scrutinize the data, especially to check if the data is consistent with assumptions required for modeling. This stage is related to exploratory data analysis (EDA), coined by @tukey with a focus on learning from data. EDA can be considered to encompass IDA. In practice, the three stages of cleaning, summarizing, and exploring are cyclical, that one often needs to do more cleaning after scrutinizing. @DasuTamraparni2003Edma say that data cleaning and exploration is a difficult task and typically consumes a large percentage of the time spent in analyzing data.

Our approach to cleaning builds heavily on the `tidyverse` approach [@tidyverse]. The data is first organized into "tidy data" [@WickhamHadley2014TD] and then further wrangled using step-wise piping with a split-apply-combine strategy for mutating new variables [@plyr]. Tidy data shouldn't be confused with "tame data", which @tamedata coined to refer to textbook data sets suitable for teaching, particularly teaching statistics. The resulting (tame) data is provided in a new R package called `r yowie`, which includes the code so that the process is reproducible and could be used to further refresh the data as new records are made available in the NLSY79 database.

@@

@SingerJudithD2003Alda used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling of wages on workforce experience, with covariates education and race. This data has been playing an important role in research in various disciplines, including but not limited to economics, sociology, education, public policy, and public health for more than a quarter of the century [@MichaelRPergamit2001DWTN]. In addition, this is considered a carefully designed longitudinal survey with high retention rates, making it suitable for life course research [@MichaelRPergamit2001DWTN; @eliznlsy]. According to @eliznlsy, thousands of articles and hundreds of book chapters and monographs have utilized this data. Moreover, the NLSY79 is considered the most widely used and most important cohort in the survey data [@MichaelRPergamit2001DWTN].  <!--AE comment: For example, I would like more discussion about the textbook selected as the exemplar. It sounds like this dataset is used in many books, why choose that one? The text should include the name of the book, so readers do not need to flip to the citations in order to learn what it was. From the name, I am guessing that the book focuses heavily on this dataset, which is why it makes sense to reproduce their subset. Be explicit about this.-->

Our aim is to refresh the wages textbook data and append it with data from 1994 through to the latest data reported in 2018, a purpose that is consistent with @grimshaw's statistics education goal of embracing authentic data experiences. Here, we investigate the process of getting from the raw NLSY79 data to a textbook data set as similar as possible to that provided by @SingerJudithD2003Alda. We should also note that race is a variable in the original data set, and for compatibility, it is also provided with the refreshed data for the *purposes of studying racism, not race* [@racismnotrace]. There are a number of data sets provided by @SingerJudithD2003Alda, and we focus only on this one because it has captivated our attention for a number of years. We use it in our own teaching of longitudinal data analysis and would very much like to use data since 1994. <!--However, we cannot create the exact same data set as published in their book since we do not have the information of what age threshold they used to determine the high school dropouts.-->

@@

We refresh this data by re-creating the full data with records from survey years 1979 through to 2018 (the most recent year published). We also modify some variables. For example, use a single categorical race variable instead of the two binary race variables. We also plan to include additional variables, some for the purpose of providing more options for data exploration in teaching examples: year of the survey, age of individual in 1979, whether the individual completed high school with a diploma or with a graduate equivalency degree (GED), the highest grade completed in the corresponding year of survey, the number of jobs that the individual had in the corresponding year of survey, the total number of hours the individual usually works per week, the year when the individual started to work, and the number of years the individual worked. We do not attempt to re-create the unemployment rate variable.

The plan is to create three datasets as follows:

1. The wages data of the whole NLSY79 cohort, including females.
2. A separate table of the demographic data of the whole NLSY79 cohort.
3. The high school dropouts' wages data is closest to a refreshed version of @SingerJudithD2003Alda's data.

@@

@LooMarkvander2018Sdcw, in the context of official statistics, describe the "statistical value chain", which includes various production stages of the data cleaning process as raw data (data in the initial that it arrives), input data (data organized with correct type and identified variables), and valid data (data that has been cleaned and more accurately represents the intent of variables). What we have colorfully named as wild data can be considered to be raw data, and valid data could be considered to be textbook data in the above statistical value chain. In this section, we outline the steps to download the raw data (Section \@ref(getdata)) and then tidy the raw data into input data, specifically for the demographic variables (Section \@ref(tidydemog)) and the employment variables (Section \@ref(tidyemp)), so that the resulting input data can be used downstream for validating the data as described in Section \@ref(ida) and Section \@ref(censor).

@@

According to @WickhamHadley2014TD, tidy data sets comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table. The raw data, `new_data_qnames`, does not comply with these rules as it is organized such that each row corresponds to an individual. As respondents can have multiple jobs at specific years, the column names, such as `HRP1_1979`, `HRP2_1979`, `HRP1_1980` and `HRP2_1980`, contain the information about the job number up to 5 (`HRP1` = job 1, `HRP2` = job 2) and the year. The raw data consequently has a large number of columns (`r ncol(new_data_qnames)` to be specific). The values in the cell under the variables that begin with `HRP` correspond to the hourly wage in dollars. A glimpse of this data shows:

@@

Thus, we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars, and the demographic variables. This is done using the `tidyverse` suite of packages [@tidyverse]: `tidyr` [@tidyr] to pivot the data into long-form, with `dplyr` [@dplyr] and `stringr` [@stringr] to mutate new variables from the downloaded data from the database, and code levels of factors by text wrangling. The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data. The code for tidying the data are demonstrated at \url{`r yowie_pkgdown`/articles/raw-to-input-data.html} but also described in the subsequent subsections.

@@

For `gender` and `race`, we only rename the column names provided in the raw data, `new_data_qnames`. It is worth noting that `gender` comes from a variable called `sex` in the database. We use the term `gender` as the reports about the data on various pages in the website more commonly use `gender`. Gender, as provided in the data, is self-reported and only has two categories, "male" and "female". The interchangeable use of these terms reflects that gender and sex are sometimes regarded as similar in surveys, although they are not. Measuring gender as a binary variable has the potential to fail to capture people who do not identify themselves as either male or female or people whose gender does not align with their sex classification [@kennedy2020using]. From a statistical perspective, this sometimes difficult the adjustment of survey statistics to the population when gender is measured, for example, with three categories, and the census measured it with only two categories and used the term sex [@kennedy2020using]. Hence, for modern societal studies, it is highly recommended to separate sex and gender in the survey question and provide more options to respond to questions on gender. This discourse should also be emphasized in teaching that using these two variables should be accompanied by special attention. Similarly, this also relates to race, as reported in the database. When doing an analysis with this variable, one should keep in mind that the purpose is to study racism rather than race.

@@

* If the hourly rate is recorded as 0, we set wage as missing.
* If the total hours of worked for the corresponding job is greater than 84 hours, we set the wage and hour worked as missing.

@@

For `stwork` variable,  we only rename the column names from `new_data_qnames`. This variable is then used to calculate the next variable, `yr_wforce` for each survey round, which is the year of survey (`year`) minus the year of individual started working (`stwork`). Finally, `exp` variable is derived from the number of weeks worked since the last interview indicated as variable started with `WKSWK` in `new_data_qnames`. To obtain the work experience since 1979, we calculate the cumulative value. As the measurement unit is in week, we convert this to year.

@@

Experience is years after entering the labor force. It represents the difference between the day an individual enters the labor force (`EMPLOYERS_ALL_STARTDATE_ORIGINAL.01~Y_XRND` in the database) relative to the date of the survey, which we call `yr_wforce`. However, using this calculation produced numbers that do not quite match with the original data.

Reading the section titled ["Topical Guide to the Data"](https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience) in the guide suggests that it should be calculated based on the variable "number of weeks worked since the last interview". This would remove periods of unemployment which makes sense when measuring experience while actually working. Since it is only measured since the last interview, this needs to be cumulated for each survey year. This produced results more similar to the original data (as discussed further in Section \@ref(takeaways)).

@@

According to @HuebnerMariannePhD2016Asat, initial data analysis (IDA) is the step of inspecting and screening the data after collection to ensure that the data is clean, valid, and ready to be deployed in the later analyses. This is supported by @Chatfield1985TIEo, who argues that the two main objectives of IDA are data description, which is to assess the structure and the quality of the data, and model formulation without any formal statistical inference.

In this paper, we conduct an IDA or a preliminary data analysis to assess the validity of the variable values in the cohort of data that the NLSY provides. <!-- AE comments: This comes as a surprise to the reader. What anomaly? The wage anomalies should be mentioned in the introduction so as not to be a surprise at this point in the manuscript.
 --> The first step is validating numerical summaries of the raw data are the same as reported by NLSY79. This is followed by graphical summaries using methods available in `ggplot2` [@ggplot2] and `brolgar` [@brolgar].

The respondents' ages ranged from 12 to 22 when first interviewed in 1979. Hence, we validate whether all of the respondents were in this range in the data we extracted. Additionally, the [NLSY](https://www.nlsinfo.org/content/cohorts/nlsy79/intro-to-the-sample/nlsy79-sample-introduction) also provides the number of the survey cohort by their gender (6,403 males and 6,283 females) and race (7,510 Non-Black/Non-Hispanic; 3,174 Black; 2,002 Hispanic). To validate this, we used the `demog_nlsy79`, i.e., the data with the survey years 1979 sample. Tables \ref{tab:age-table} and \ref{tab:gender-race-table} suggest that the demographic data we had is consistent with the sample information in the database.

@@

<!--AE comment: With the IDA about wage data, I would appreciate a bit more context. Why sample 36 observations? I am not a longitudinal data expert, so I was a bit surprised when I flipped to the plot and found it to be small-multiple time series, although upon further consideration it makes sense. In that section it would be good to describe what the figure shows in a bit more detail. Perhaps “We randomly sample 36 respondents from the data, and plot their average wage as a time series. Looking at the patterns over time, we see a lot of variability in wages. For example, the people shown in panels 5, 7, and 11 have [explain what is interesting here, again I’m not an expert].” The next sentence here, “Some have had flat wages for years but had a sudden increase in one particular year, then it gone down again, while the others experienced an upsurge in their wage, for instance, the IDs in panel 9.” does not seem to provide much additional information. What were you looking for here? Overall trends over time, increasing/decreasing? Any places where one year’s wage looked really different?-->

In the next step, we explore the mean hourly wage data of samples of individuals. The purpose is to examine the common patterns and check the quality. A random sample of 36 individuals is chosen (using the `sample_n_keys` function in `brolgar`). Their longitudinal profiles are plotted, faceted by `id`, and using free $y$ scales so that the individual patterns can be examined (Figure \ref{fig:sample-plot}).<!--AE comment: "It" implies... Please specify what "it" refers to. This could be done by adding onto the first sentence... "as shown in Figure 2, which shows that some respondents have high variability in wages, particularly 5, 7, and 11." However, 7 and 11 don't seem to have particularly high variability, especially when compared to 18. This whole paragraph needs to be fixed/updated.--> There is a lot of variability from one individual to another and substantial fluctuation in wages at different times for most individuals. Some individuals  (2799, 11041, 11146) are only measured for a short period. Some individuals (8296, 9962) possibly have errors in wages in some years because of the extreme fluctuation. These need to be inspected more closely. It is also important to note that some shorter profiles indicate that some individuals have left the study before it has finished. Checking whether the demographics of the early departing are similar to those who remain in the study is an important part of any downstream analysis to account for the bias induced by the inadvertent censoring.

 <!--AE comment: Not panel 9  the IDs in panel 9. -->
<!--I think there is a narrative step missing here, which is that looking at those samples led you to look at plots that show values for the wage variable over all participants and all years. The phrase “the summary plots” was not enough description for me to know what to expect when I flipped to Figure 3. Please be explicit about what the three plots were, why they were made, and what you were looking for. I think there is another narrative step missing about Figure 3C, which is that after you saw there were outliers you tried to identify which respondents had the extreme values. Again, please spell this out in the text. When you return to the plots for the entire dataset in Figure 5, please make parallel comments. -->

Figure \ref{fig:feature-plot}s shows an alternative way to check the data quality. Plot (A) is the spaghetti plot where all profiles are shown, and it can be seen that there are unbelievably high wage values (up to $60,000/hour) for some individuals, mostly around 1990. Plot (B) shows side-by-side boxplots of the three number summaries (minimum, median, and maximum) for all individuals. This tells us that there are a number of individuals with unbelievably high maximum wages. Plot (C) shows the profile for an individual, with not such a high maximum wage but still indicates a problem: their wages are consistently low except for one year where they earned close to $1200/hour. This does not seem to be reasonable and leads us to use a procedure to detect and fix these temporal anomalies.

@@

<!--AE comment: Table 2 seems to have more precision than is necessary. Certainly 100% does not need to be reported as 100.00%, and probably all the percentages could be rounded to the nearest 1 percent. The caption could explain this, as there might be some rounding errors. -->

@@

<!--AE comment: Figure 2’s caption should include something about how that data is the raw values from the survey, before inputing anomalous values so the figure and caption can stand alone.-->

<!--For both Figure 3 and Figure 5, I would prefer the letter label before the description, but that’s a matter of personal preference.-->

```{r feature-plot, fig.cap = "Summary plots to check the data after the tidying stage: (A) longitudinal profiles of wages for all individuals 1979-2018, (B) boxplots of minimum, median, and maximum wages of each individual, (C) and one individual (id=39) with an unusual wage relative to their years of data. It reveals that some values of hourly wages are unbelievable, and some individuals have extemely unusual wages in some years. Accordingly, more cleaning is necessary to treat these extreme values.", fig.width=10, fig.height=4, out.width="100%", cache = FALSE}
```

@@

Figure \ref{fig:compare-plot} shows the mean hourly wage before and after the extreme values are replaced. The plot shows that fluctuations in wages remain, but the large spikes (in this sample, individuals 8296, 9962), which are considered implausible, are replaced.

>>>>

---
title: "A Journey from Wild to Textbook Data to Reproducibly Refresh the Wages Data from the National Longitudinal Survey of Youth Database"
authors:
  - name: Dewi Amaliah
    thanks: ""
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: dlamaleeah@gmail.com
  - name: Dianne Cook
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: dicook@monash.edu
  - name: Emi Tanaka
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: emi.tanaka@monash.edu
  - name: Kate Hyde
    thanks: ""
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: "hyde.kate.a@gmail.com"
  - name: Nicholas Tierney
    thanks: ""
    department: ""
    affiliation: Telethon Kids Institute
    location: Nedlands, WA 6009
    email: "nicholas.tierney@gmail.com"
bibliography: references.bib
biblio-style: unsrt
#preamble: >
header-includes:
  - \usepackage{tcolorbox}
  - \usepackage{fontawesome}
  - \usepackage{color, colortbl}
output: 
  #rticles::arxiv_article:
    #keep_tex: false
  bookdown::pdf_book:
    base_format: rticles::arxiv_article
date: "24/02/2021"
always_allow_html: true
keywords: "Data cleaning; Data tidying; Reproducible workflow; Longitudinal data; NLSY79; Initial data analysis;"
abstract: > 
 Textbook data is essential for teaching statistics and data science methods because they are clean, allowing the instructor to  focus on methodology. Ideally textbook data sets are refreshed regularly, especially when they are subsets taken from an on-going data collection. It is also important to use contemporary data for teaching, to imbue the sense that the methodology is relevant today. This paper describes the trials and tribulations of refreshing a textbook data set on wages, extracted from the National Longitudinal Survey of Youth (NLSY79) in the early 1990s. The data is useful for teaching modeling and exploratory analysis of longitudinal data. Subsets of NLSY79, including the wages data, can be found in supplementary files from numerous textbooks and research articles. The NLSY79 database has been continuously updated through to 2018, so new records are available. Here we describe our journey to re-create the wages data, and document the process so that the data can be regularly updated into the future. Our journey was difficult because the steps and decisions taken to get from the raw data to the wages textbook subset have not been clearly articulated. We have been diligent to provide a reproducible workflow for others to follow, which also hopefully inspires more attempts at refreshing data for teaching. Three new data sets and the code to produce them are provided in the open source R package called [CENSORED]. 
---

@@

Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques. An example of this is the wages data made public by @SingerJudithD2003Alda in their book, "Applied longitudinal data analysis," which can be used to teach generalized linear models, in addition to hierarchical, mixed effects, and multilevel models. The data records hourly wages of a sample of high school dropouts from 1979-1994, along with the demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) [@nlsy79].

The story from modeling the data (and as reported by Singer and Willett) is that wages increase with the length of time in the workforce, a higher level of education leads to higher wages, and that race makes a difference, on average. However, an exploratory analysis reveals that the individual experience varies a great deal from the overall average. Some individuals experience a decline in wages the longer they are in the workforce, and many experience volatility in their wages. The wages data was used to illustrate exploratory longitudinal data analysis in @ilk2004, and was further developed into a case study for use in the teaching of exploratory data analysis at `r censor_word("Iowa State University")`.

@@

This disparity between the average and the individual is a part of statistics as a discipline that requires more attention. This particular data set is a prime example of discussing this disparity. Textbook data sets have longevity if there is an unresolved mystery. The iris data [@iris-data] is a prime example. It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis (a side note: the iris data is best replaced today with the penguins data [@penguins-data], which has similar qualities, is new, and does not suffer from a connection with eugenics [@notiris]). We argue the wages data is in this class of textbook data, too, because it presents a challenge for longitudinal data analysis: how can we better summarize and explain the individual experience?

For statistics and data science by association, it is increasingly important to reach the individual. One might describe this as a divergence of purpose -- statistics for public policy -- or statistics for the public. The two are not the same. As the world becomes more electronically connected, combating misinformation and mitigating conspiracy theories require statistics that address the individual. For example, with the wages data, even though the message for public policy is that demographic profile is related to different wage patterns on average, the message for the individual is you are more than your demographic. The majority of people in the study do not have a pattern that is similar to the average. If you have a bad experience, that your wages have declined over time, you are not alone; there are others like you, and more than you think. A similar tone is echoed occasionally in the public media. For example, an article published in the Sydney Morning Herald argues there is no average Australian [@notaverage].

@@

This paper demonstrates the steps of cleaning data, including subjective decisions made on dealing with anomalies, and documents the process, as recommended by @HuebnerMariannePhD2016Asat. They emphasize that making the data cleaning process accountable and transparent is imperative and essential for the integrity of downstream statistical analyses and model building. Clean data often then goes through an "initial data analysis" (IDA) [@Chatfield1985TIEo], where one would summarize and scrutinize the data, especially to check if the data is consistent with assumptions required for modeling. This stage is related to exploratory data analysis (EDA), coined by @tukey with a focus on learning from data. EDA can be considered to encompass IDA. In practice, the three stages of cleaning, summarizing, and exploring are cyclical, one often needs to do more cleaning after scrutinizing. @DasuTamraparni2003Edma say that data cleaning and exploration is a difficult task and typically consumes a large percentage of the time spent in analyzing data.

Our approach to cleaning builds heavily on the `tidyverse` approach [@tidyverse]. The data is first organized into "tidy data" [@WickhamHadley2014TD] and then further wrangled using step-wise piping with a split-apply-combine strategy for mutating new variables [@plyr]. Tidy data shouldn't be confused with "tame data," which @tamedata coined to refer to textbook data sets suitable for teaching, particularly teaching statistics. The resulting (tame) data is provided in a new R package called `r yowie`, which includes the code so that the process is reproducible and could be used to further refresh the data as new records are made available in the NLSY79 database.

@@

@SingerJudithD2003Alda used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling of wages on workforce experience, with covariates education and race. This data has been playing an important role in research in various disciplines, including but not limited to economics, sociology, education, public policy, and public health for more than a quarter of the century [@MichaelRPergamit2001DWTN]. In addition, this is considered a carefully designed longitudinal survey with high retention rates, making it suitable for life course research [@MichaelRPergamit2001DWTN; @eliznlsy]. According to @eliznlsy, thousands of articles and hundreds of book chapters and monographs have utilized this data. Moreover, the NLSY79 is considered the most widely used and most important cohort in the survey data [@MichaelRPergamit2001DWTN].

Our aim is to refresh the wages textbook data and append it with data from 1994 through to the latest data reported in 2018, a purpose consistent with @grimshaw's statistics education goal of embracing authentic data experiences. Here, we investigate the process of getting from the raw NLSY79 data to a textbook data set as similar as possible to that provided by @SingerJudithD2003Alda. We should also note that race is a variable in the original data set, and for compatibility, it is also provided with the refreshed data for the *purposes of studying racism, not race* [@racismnotrace]. There are a number of data sets provided by @SingerJudithD2003Alda, and we focus only on this one because it has captivated our attention for a number of years. We use it in our own teaching of longitudinal data analysis and would very much like to use data since 1994. <!--However, we cannot create the exact same data set as published in their book since we do not have the information of what age threshold they used to determine the high school dropouts.-->

@@

We refresh this data by re-creating the full data with records from survey years 1979 through to 2018 (the most recent year published). We also modify some variables. For example, we use a single categorical race variable instead of the two binary race variables. We also include additional variables, some for the purpose of providing more options for data exploration in teaching examples: year of the survey, age of individual in 1979, whether the individual completed high school with a diploma or with a graduate equivalency degree (GED), the highest grade completed in the corresponding year of survey, the number of jobs the individual had in the corresponding year of survey, the total number of hours the individual usually works per week, the year when the individual started to work, and the number of years the individual worked. We do not attempt to re-create the unemployment rate variable.

We aim to create three datasets as follows:

1. The wages data of the whole NLSY79 cohort, including females.
2. A separate table of the demographic data of the whole NLSY79 cohort.
3. The subset of wages data in (1), which is the wages of high school dropouts' as a refreshed version of @SingerJudithD2003Alda's data.

@@

In the context of official statistics, @LooMarkvander2018Sdcw describe the "statistical value chain," which includes various production stages of the data cleaning process as raw data (data in the initial that it arrives), input data (data organized with correct type and identified variables), and valid data (data that has been cleaned and more accurately represents the intent of variables). What we have colorfully named wild data can be considered to be raw data, and valid data could be considered to be textbook data in the above statistical value chain. In this section, we outline the steps to download the raw data (Section \@ref(getdata)) and then tidy the raw data into input data, specifically for the demographic variables (Section \@ref(tidydemog)) and the employment variables (Section \@ref(tidyemp)), so that the resulting input data can be used downstream for validating the data as described in Section \@ref(ida) and Section \@ref(censor).

@@

According to @WickhamHadley2014TD, tidy data sets comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table. The raw data, `new_data_qnames`, does not comply with these rules as it is organized such that each row corresponds to an individual. As respondents can have multiple jobs at specific years, the column names, such as `HRP1_1979`, `HRP2_1979`, `HRP1_1980` and `HRP2_1980`, contain the information about the job number up to 5 (`HRP1` = job 1, `HRP2` = job 2) and the year. The raw data consequently has a large number of columns (`r ncol(new_data_qnames)` to be specific). The values in the cell under the variables begin with `HRP` correspond to the hourly wage in dollars. A glimpse of this data shows:

@@

Thus, we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars, and the demographic variables. This is done using the `tidyverse` suite of packages [@tidyverse]: `tidyr` [@tidyr] to pivot the data into long-form, with `dplyr` [@dplyr] and `stringr` [@stringr] to create new variables from the downloaded data from the database, and code levels of factors by text wrangling. The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data. The code for tidying the data are demonstrated at \url{`r yowie_pkgdown`/articles/raw-to-input-data.html} but also described in the subsequent subsections.

@@

For `gender` and `race`, we only rename the column names provided in the raw data, `new_data_qnames`. It is worth noting that `gender` comes from a variable called `sex` in the database. We use the term `gender` as the reports about the data on various pages in the website more commonly use `gender`. Gender, as provided in the data, is self-reported and only has two categories, "male" and "female." The interchangeable use of these terms reflects that gender and sex are sometimes regarded as similar in surveys, although they are not. Measuring gender as a binary variable has the potential to fail to capture people who do not identify themselves as either male or female or people whose gender does not align with their sex classification [@kennedy2020using]. From a statistical perspective, this sometimes difficult the adjustments of survey statistics to the population when gender is measured, for example, with three categories, and the census measured it with only two categories and used the term sex [@kennedy2020using]. Hence, for modern societal studies, it is highly recommended to separate sex and gender in the survey question and provide more options to respond to questions on gender. This discourse should also be emphasized in teaching that using these two variables should be accompanied by special attention. Similarly, this also relates to race, as reported in the database. When doing an analysis with this variable, one should keep in mind that the purpose is to study racism rather than race.

@@

* If the hourly rate is recorded as 0, we set wage as missing.
* If the total hours of worked for the corresponding job is greater than 84 hours, we set the wage and hours worked as missing.

@@

For `stwork` variable,  we only rename the column names from `new_data_qnames`. This variable is then used to calculate the next variable, `yr_wforce` for each survey round, which is the year of survey (`year`) minus the year of individual started working (`stwork`). Finally, `exp` variable is derived from the number of weeks worked since the last interview indicated as variable started with `WKSWK` in `new_data_qnames`. To obtain the work experience since 1979, we calculate the cumulative value. As the measurement unit is weeks, we convert this to year.

@@

Experience is years after entering the labor force. It represents the difference between the day an individual enters the labor force (`EMPLOYERS_ALL_STARTDATE_ORIGINAL.01~Y_XRND` in the database) relative to the date of the survey, which we call `yr_wforce`. However, using this calculation produced numbers that do not quite match the original data.

Reading the section titled ["Topical Guide to the Data"](https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience) in the guide suggests it should be calculated based on the variable "number of weeks worked since the last interview." This would remove periods of unemployment which makes sense when measuring experience while actually working. Since it is only measured since the last interview, this needs to be cumulated for each survey year. This produced results more similar to the original data (as discussed further in Section \@ref(takeaways)).

@@

According to @HuebnerMariannePhD2016Asat, initial data analysis (IDA) is the step of inspecting and screening the data after collection to ensure the data is clean, valid, and ready to be deployed in the later analyses. This is supported by @Chatfield1985TIEo, who argues that the two main objectives of IDA are data description, which is to assess the structure and the quality of the data, and model formulation without any formal statistical inference.

In this paper, we conduct an IDA or a preliminary data analysis to assess the validity of the variable values in the cohort of data that the NLSY provides. The first step is validating numerical summaries of the raw data are the same as reported by NLSY79. This is followed by graphical summaries using methods available in `ggplot2` [@ggplot2] and `brolgar` [@brolgar].

The respondents' ages ranged from 12 to 22 when first interviewed in 1979. Hence, we validate whether all respondents were in this range in the data we extracted. Additionally, the [NLSY](https://www.nlsinfo.org/content/cohorts/nlsy79/intro-to-the-sample/nlsy79-sample-introduction) also provides the number of the survey cohort by their gender (6,403 males and 6,283 females) and race (7,510 Non-Black/Non-Hispanic; 3,174 Black; 2,002 Hispanic). To validate this, we used the `demog_nlsy79`, i.e., the data with the survey years 1979 sample. Tables \ref{tab:age-table} and \ref{tab:gender-race-table} suggest the demographic data we had is consistent with the sample information in the database.

@@

In the next step, we explore the mean hourly wage data of samples of individuals. The purpose is to examine the common patterns and check the quality. A random sample of 36 individuals is chosen (using the `sample_n_keys` function in `brolgar`). Their longitudinal profiles are plotted, faceted by `id`, and using free $y$ scales so that the individual patterns can be examined (Figure \ref{fig:sample-plot}). There is a lot of variability from one individual to another and substantial fluctuation in wages at different times for most individuals. Some individuals  (2799, 11041, 11146) are only measured for a short period. Some individuals (8296, 9962) possibly have errors in wages in some years because of the extreme fluctuation. These need to be inspected more closely. It is also important to note that some shorter profiles indicate some individuals have left the study before it has finished. Checking whether the demographics of the early departing are similar to those who remain in the study is an important part of any downstream analysis to account for the bias induced by the inadvertent censoring.

Figure \ref{fig:feature-plot} shows an alternative way to check the data quality. Plot (A) is the spaghetti plot where all profiles are shown, and it can be seen that there are unbelievably high wage values (up to $60,000/hour) for some individuals, mostly around 1990. Plot (B) shows side-by-side boxplots of the three number summaries (minimum, median, and maximum) for all individuals. This tells us there are a number of individuals with unbelievably high maximum wages. Plot (C) shows the profile for an individual with a maximum wage that is not so extreme but still indicates a problem: their wages are consistently low except for one year where they earned close to $1200/hour. This does not seem to be reasonable and leads us to use a procedure to detect and fix these temporal anomalies.

@@

@@

```{r feature-plot, fig.cap = "Summary plots to check the data after the tidying stage: (A) longitudinal profiles of wages for all individuals 1979-2018, (B) boxplots of minimum, median, and maximum wages of each individual, (C) and one individual (id=39) with an unusual wage relative to their years of data. It reveals that some values of hourly wages are unbelievable, and some individuals have extremely unusual wages in some years. Accordingly, more cleaning is necessary to treat these extreme values.", fig.width=10, fig.height=4, out.width="100%", cache = FALSE}
```

@@

Figure \ref{fig:compare-plot} shows the mean hourly wage before and after the extreme values are replaced. The plot shows fluctuations in wages remain, but the large spikes (in this sample, individuals 8296, 9962), which are considered implausible, are replaced.
